---
title: "CONTRACEPTIVE USE IN MEDICAL CONTEXT: QUANTITATIVE REPORT"
output:
  prettydoc::html_pretty:
    theme: architect
    toc: true
    self_contained: TRUE
author: "Léo Persoz (DNDi, UNIGE)"
date: "December 2024"
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

```{r install, include=FALSE}
# install/load packages, load dataframe(s)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse,
               readxl,
               plotly,
               reactable,
               gt,
               gtsummary,
               gtExtras,
               here)

graphs_dataset <- read_csv(here("C:/Kenya_anaylsis_Rproject/Project_outputs/data/graphs_dataset.csv"))
```

# 2. Methodology

### 2.1 Study design and period

This study consists in the quantitative phase of a mixed methods
research project on the topic of medically indicated contraceptive use.
The project relies on a explanatory sequential design, and its
quantitative strands follows a hospital-based cross-sectional design
implemented through a structured survey. It was administered between the
21 and 25 October 2024, and collects information on the topic of
intention to use contraceptives when indicated alongside medical
treatment for leishmaniasis. Before being triangulated with the
qualitative results, the findings presented in this report will be used
in order to refine tools related to the qualitative phase of the study
(i.e. in depth interview guide and focus-group discussion guide), as
well as inform the sampling of in depth interview participants.

### 2.2 Study area

The questionnaire was administered in the only two health centers
delivering treatment for leishmaniasis situated in West Pokot, Kenya:
Kacheliba and Sigor District Hospitals.

### 2.3 Inclusion and exclusion criteria

#### Inclusion criteria

-   All women of reproductive age frequenting the Kacheliba or Sigor
    District Hospitals

### 2.4 Sample size calculation and sampling procedure

### 2.5 Data collection instrument

### 2.6 Data processing and analysis

### 2.7 Quality control

### 2.8 Operational definitions

-   Economic status

-   Knowledge of Kala-azar

-   Experience of Kala-azar

-   N of consequences feared

-   contraceptive knowledge

    -   0-1 low

    -   2-4 moderate

    -  > 5 high

-   fear of side effects

    -   total of side effect feared divided by the total of side effects
        known, varying between 0 and 1.

        -   0-0.33 low fear

        -   0.34-0.66 moderate fear

        -   0.67-1 high fear

-   CU approval

-   CU agency

-   Decision maker

# 3. Results

## Descriptive analysis

A total of 401 WOCBP were interviewed from Kacheliba and Sigor District
Hospitals. Socio-demographic characteristics are found in Table 1
(numeric variables) and Table 2 (categorical variables).

##### Table 1

```{r Table 1, echo=FALSE}
htmltools::tags$iframe(
  src = here::here("C:/Kenya_anaylsis_Rproject/Project_outputs/outputs_touse/tables/final_num_vars_table.html"),
  width = "100%",
  height = "250px",
  style = "border:none;"
  )
```

Concerning the age of participants, the distribution is a little
negatively skewed, due to a higher number of participants of young age.
Participants also show a similar number of birth(s) and number of
children they are living with.

##### Table 2

```{r Table 2, echo=FALSE}
htmltools::tags$iframe(
  src = here::here("C:/Kenya_anaylsis_Rproject/Project_outputs/outputs_touse/tables/final_cat_var_table.html"),
  width = "100%",
  height = "600px",
  style = "border:none;"
  )
```

#### *Socio-demographic variables*

A majority of participants are living in rural areas (92.3%), sedentary
(65.8%), and of Christian confession (96.8%). Furthermore, Almost half
participants did not complete any level of formal education (48.1%) ,
and a quarter did complete primary education (25.7%). Most participants
were considered as having a low economic status (81%) or medium economic
status (18.2%).

#### *Knowledge and perceptions of leishmaniasis*

Knowledge of leishmaniasis was quite high among participants, with 33.4%
of the sample knowing both at least a mean of transmission and one
symptom, and 29.4% of participants who either never heard about
leishmaniasis, or did not know any symptom nor means of transmission.
Most participants had some experience with leishmaniasis, with 56.6% of
the sample having heard or seen someone sick from leishmaniasis, and
21.7% heard or seen someone die from this disease. Furhtermore,
participants showed relatively low concerns about consequences of
leishmaniasis treatment, with above 70% of participants evaluated as
having no or low concerns. Pain at the injection site (31.7%) and
prolonged hospital stay (26.9%) were the most frequent concerns
mentioned by participants.

#### *Contraceptive use variables*

Approximately a third of participants reported using contraceptives when
interviewed (34.7%), while this rate jumps to 55.9% when reporting about
ever usage of contraceptives. Figure 1 describes the kind of
contraceptives methods reported to be used by participants who ever used
contraceptives. Three quarter of participants reported being the
decision makers related their use of contraceptives (75.1%), and 10.7%
reported to have been pressured to give birth at some point in the past.

##### Figure 1


```{r Figure 1, echo=FALSE}
N_ever_user <- sum(!is.na(graphs_dataset$use_type))

univar_use_type <- 
  graphs_dataset %>% 
  mutate(use_type = case_when(use_type == 1 ~ "Female sterilization",
                              use_type == 2 ~ "Male sterilization",
                              use_type == 3 ~ "Intra-uterine device (IUD)",
                              use_type == 4 ~ "Injectables",
                              use_type == 5 ~ "Implants",
                              use_type == 6 ~ "Pills",
                              use_type == 7 ~ "Condom",
                              use_type == 8 ~ "Female condom",
                              use_type == 9 ~ "Emergency contraception",
                              use_type == 10 ~ "Standard days method",
                              use_type == 11 ~ "Lactational amenorrhea method",
                              use_type == 12 ~ "Other")) %>%
  count(use_type) %>% 
  mutate(percentage = n / N_ever_user * 100,
         tooltip_label = paste(use_type, " (n = ", n, "): ", round(percentage, 2), "%", sep = "")) %>%
  filter(!is.na(use_type)) %>% 
  ggplot(aes(x = reorder(use_type, n), y = percentage, fill = use_type, text = tooltip_label)) +
  geom_col() +
  labs(title = "Type of modern contraceptive methods used among ever-users",
       x = "Type of modern contraceptive method",
       y = "Percentage") +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plotly_univar_use_type <- ggplotly(univar_use_type, tooltip = "text")

plotly_univar_use_type
```

More than three quarter of participants reported being strongly likely
(48.1%) or likely (30.7%) to use contraceptives for family planning in
the future. In the context of clinical trials, 81% of participants
reported being either strongly likely (50.1%) or likely (30.9%) to
choose a treatment regiment that requires to use a contraceptive method
during treatment over one that do not require contraceptive use,
provided that the former treatment is less disruptive (in terms of
number of injections, length of hospital stay and side effects) - this
rate jumps to 85.1% (strongly likely: 58.4%, likely: 26.7%) in the
context of standard treatment (i.e. outside clinical trials).

#### *Psycho-social variables*

Participants mostly displayed moderate contraceptive knowledge (72.6%),
and had high fear of side effects(45.4%). The kind of feared side
effects are displayed in Figure 2. Concerning contraceptive use
approval, three quarter of the sample displayed high approval (75. 8%) -
however, 45.6% of participants described a high agency regarding
contraceptive use, and 42.4% were categorized as having moderate agency.

##### Figure 2

```{r Figure 2, echo=FALSE}
univar_fear_side_eff <- 
  graphs_dataset %>%
  # Split combined_reasons into individual reasons, while retaining the original structure
  mutate(f_side_eff_combined_list = str_split(f_side_eff_combined, ",\\s*")) %>%
  unnest(f_side_eff_combined_list) %>%  # Create a column of individual reasons
  filter(!is.na(f_side_eff_combined_list)) %>%  
  count(f_side_eff_combined_list) %>% 
  mutate(percentage = n / sum(n) * 100, # nrow(num_data_renamed) inserted manually to avoid counting 
         # from split reasons rather than the total sample size
         tooltip = paste(f_side_eff_combined_list, " (n = ", n, ") ", round(percentage, 2), "%", sep = "")) %>% 
  ggplot(aes(x = reorder(f_side_eff_combined_list, n), y = n, fill = f_side_eff_combined_list, text = tooltip)) +
  geom_col() +
  labs(x = "Side effects feared", y = "Count", title = "Counts of side effects feared by respondents") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none")

plotly_univar_fear_side_eff <- ggplotly(univar_fear_side_eff, tooltip = "text")

plotly_univar_fear_side_eff
```

## Binary analysis: difference in proportions

In order to establish whether a significant difference exists between
participants reporting future intentions to use contraceptives for
family planning and intentions to use some contraceptive methods in the
context of standard treatment for leishmaniasis, a McNemar test was
applied these two proportions. To do so, both of these variables were
binarized - as shown in Table 2, both were originally recorded using a 4
point likert-scale. The results of this test are found in Table 3.

##### Table 3

```{r echo=FALSE}
htmltools::tags$iframe(
  src = here::here(
    "C:/Kenya_anaylsis_Rproject/Project_outputs/outputs_touse/tables/final_mcnemar_gt_treat.html"),
  width = "100%",
  height = "325px",
  style = "border:none;"
  )
```


## Multivariate analysis

Table 4-6 display three logistical regression models that differ
structurally because of their respective outcomes: Table 4 includes
intentions to use contraceptives alongside medical treatment as outcome,
Table 5 includes future intentions to use contraceptives as outcome,
while Table 6 includes ever use of contraceptive as outcome. These
models have been set up in order to compare the difference in predictors
for each outcome, thereby allowing to identify difference in factors
influencing contraceptives use intentions (and behaviour for ever use)
in these different contexts (for medical reasons and for family planning
purpose).

##### Table 4

```{r echo=FALSE}
htmltools::tags$iframe(
  src = here::here("C:/Kenya_anaylsis_Rproject/Project_outputs/outputs_touse/tables/gt_AORs_treatment_ord2.html"),
  width = "100%",
  height = "600px",
  style = "border:none;"
  )
```

Concerning the psychosocial predictors included in the model described
in Table 4, all categories related to the approval of contraceptive use
and agency with regard to contraceptive use are highly negatively
associated with intention to use contraceptives alongside medical
treatment (all p-values under 0.01) - fear of side effects seems to be
positively associated with the outcome (although in a less significant
manner as shown by the p-value), although only when observing the
moderate category. Similarly, ever use of contraceptives is also
positively associated with the outcome. However, it is important to note
that the ability of participants to decide using contraceptives for
herself is not associated with intention to use contraceptives alongside
medical treatment, as shown by the odds ration and p-value related to
this variable. Furthermore, knowledge of contraceptive shows no
association with intention to use contraceptives alongside medical
treatment.

For demographic predictors, age and number of children are both highly
associated with intention to use contraceptive alongside medical
treatment, although in different directions (negatively and positively,
respectively). Somehow surprisingly, higher economic status tends to be
negatively associated with the outcome, when one could expect it to be
positively associated.

Finally, disease related predictors also show some interesting results:
as the number of concerns related to treatment consequences, intentions
to use contraceptives for medical reasons increase. This is not
surprising, since the treatment regiment for leishmaniasis which does
involve using contraceptives present more negative consequences than the
regiment requiring to use contraceptives (a precision which was stated
in the prompt of the question related to intention to use contraceptives
alongside medical treatment). Finally, being knowledgeable about the
symptoms of leishmaniasis, in comparison to being knowledgeable about
both symptoms and means of transmission of leishmaniasis, is positively
associated with the outcome.

##### Table 5

```{r Table 5, echo=FALSE}
htmltools::tags$iframe(
  src = here::here("C:/Kenya_anaylsis_Rproject/Project_outputs/outputs_touse/tables/gt_AORs_intentionFP_ord.html"),
  width = "100%",
  height = "600px",
  style = "border:none;"
  )
```

##### Table 6

```{r Table 6, echo=FALSE}
htmltools::tags$iframe(
  src = here::here("C:/Kenya_anaylsis_Rproject/Project_outputs/outputs_touse/tables/gt_AORs_ever_ord.html"),
  width = "100%",
  height = "600px",
  style = "border:none;"
  )
```

What can be observed when comparing these three models is that first

# Discussion

-   difference between the models

    -   agency and decision maker varies (decision_maker not significant
        in medical context but significant for FP - agency strongly
        associated in medical context but not really in FP context)

    -   age and n_birth are highly significant in medical context but
        not so much for family planning

    -   "Classic" predictors of contraceptive use are not so much
        associated with the medical context, in comparison to FP
        context.

-   why age (and n_birth) is significant only in medical context

-   issue with the fear of SE variable

-   8% of the sample fears issues related to fertility when using FP

-   drinking unboiled milk have been often reported as a cause of
    leishmaniasis

What can be taken out of this table is first that a strong and
significant difference exists between the proportion compared (McNemar
Chi-Squared = 8.6, p-value = 0.0032). The absolute difference further
confirm the direction of this difference, with future intentions to use
contraceptive for family planning being 6.5% lower than intention to use
contraceptives alongside leishmaniasis treatment (CI = -10.9%, - 2.2%).
This tends to indicate that decision related to using contraceptives in
these two different contexts are actually taken differently, and
identifying predictors related to these two phenomenon are further
inquired in the subsequent section.
