---
title: "CONTRACEPTIVE USE IN MEDICAL CONTEXT: QUANTITATIVE REPORT"
output:
  prettydoc::html_pretty:
    theme: architect
    toc: true
    self_contained: TRUE
author: "Léo Persoz (DNDi, UNIGE)"
date: "December 2024"
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

```{r install, include=FALSE}
# install/load packages, load dataframe(s)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse,
               plotly,
               here)

graphs_dataset <- read_csv(here("C:/Kenya_anaylsis_Rproject/Project_outputs/data/graphs_dataset.csv"))
```

# 2. Methodology

### 2.1 Study design and period

This study consists in the quantitative phase of a mixed methods
research project on the topic of medically indicated contraceptive use.
The project relies on a explanatory sequential design, and its
quantitative strands follows a hospital-based cross-sectional design
implemented through a structured survey. It was administered between the
21 and 25 October 2024, and collects information on the topic of
intention to use contraceptives when indicated alongside medical
treatment for leishmaniasis. Before being triangulated with the
qualitative results, the findings presented in this report will be used
in order to refine tools related to the qualitative phase of the study
(i.e. in depth interview guide and focus-group discussion guide), as
well as inform the sampling of in depth interview participants.

### 2.2 Study area

The questionnaire was administered in the only two health centers
delivering treatment for leishmaniasis situated in West Pokot, Kenya:
Kacheliba and Sigor District Hospitals.

### 2.3 Inclusion and exclusion criteria

**Inclusion criteria**

-   WOCBP aged 15-49 years old living in endemic area covered by the
    Kacheliba Hospital or Sigor Sub-County Hospital, irrespective of
    their contraceptive use and/or pregnancy status.

**Exclusion criteria**

-   Women who, in the view of their treating physician (or the
    investigator if medically qualified), are not medically fit to
    participate.

### 2.4 Sample size calculation and sampling procedure

The required sample size was estimated using a single population
proportion formula, which result was adjusted using a finite population
correction to account for the population size.

As no quantitative data about contraceptive use acceptability in the
context of medical treatment have been published in this region, an
estimated proportion of 50% for contraceptive use acceptability when
recommended for medical reasons was used to maximize variance.
Confidence interval and the margin of error were set at 95% and 5%
respectively. These calculations yield a sample size of 384
participants, to which a finite population correction has been applied
based on the Kacheliba and Sigor catchment populations, which
respectively are 18,090 and 14,172 (facilities data, unpublished). For
both figures, 21.4% was taken to estimate the sampling frame, which
corresponds to the proportion of women of childbearing potential in the
catchments, yielding a result of 6904. This correction resulted in an
adjusted sample size of 364, to which a 10% non-response rate was added,
resulting in final sample size of 401 participants. The formula
mentioned here in the above are the following:

*Single population proportion formula*

$x = \frac{Z_{\frac{\alpha}{2}}^2 \cdot p \cdot (1-p)}{\text{MOE}^2}$

*Finite population correction formula*

$n = \frac{N \cdot x}{x + N - 1}$

where *Z~α/2~* is the critical value of the Normal distribution at α/2
(i.e. confidence interval), *p* is the estimated sample proportion,
*MOE* is the margin of error and *N* is the sampling frame.

### 2.5 Data collection instrument

A structured and pre-tested interview-based questionnaire was used to
collect data, which includes closed-ended and open-ended questions. The
questionnaire was administered on tablets by research assistants using
the Kobo collect app. This instrument is available in Appendix I.

### 2.6 Data processing and analysis

Data were entered, cleaned, coded and analysed using *Rstudio*, version
2024.04.02 Build 764. Descriptive and analytic statistical procedures
were applied.

Analytic statistical procedures consisted firstly in a McNemar test of
independence for paired proportions to examine independence between
intentions to use contraceptives in the context of medical treatment
(i.e. alongside medical treatment) and for family planning. Secondly,
three models were created for the bivariate and multivariate logistic
regression, including three different outcomes: intentions to use
contraceptives when medically indicated alongside Kala-azar treatment,
future intentions to use contraceptives for family planning, and ever
usage of contraceptives. For each models, a forward stepwise selection
based on the Akaike Information Criterion (AIC) was utilized to
determine the best set of predictors. Finally, likelihood ratio tests
were then performed to confirm the superiority of the final models over
simpler ones.

### 2.7 Quality control

Co-investigators introduced the data collection instrument to research
assistants (RAs), and trained them to use them. RAs were given time to
familiarize themselves with the instrument by performing role plays
among temselves, and subsequently during the piloting of the tool.
During data collection, daily meeting took place to provide a space for
RAs to communicate to their peers and the co-investigators about their
daily experience in administering the questionnaire, and voice any
questions or issues encountered during the day.

### 2.8 Operational definitions

The instrument use to collect data analysed in the subsequent section is
availble in appendix I. The description of variables which have
constructed is found just below.

-   **CU Intention alongside treatment**

    Intentions to use contraceptive alongside medical treatment for
    Kala-azar. A short prompt was provided to inform participants about
    a potential situation where she could asked if she would be willing
    to use contraceptives in this context (see Appendix I for the
    prompt). This variable is evaluated using 4 modalities, in the form
    of a likert scale (i.e. strongly likely, likely, unlikely, strongly
    unlikely). To conduct a logistical regression using this variable as
    outcome, this variable has been binarized as follows:

    -   "Some intentions": participants who answered "Strongly likely"
        or "Likely" to the CU intention alongside treatment variable.

    -   "No intention": participants who answered "Unlikely" or
        "Strongly unlikely" to the CU intention alongside treatment
        variable.

-   **CU intention during clinical trials**

    Similar question to the previous one, but in the context of a
    clinical trial and not during standard treatment. As for standard
    treatment, a prompt was provided to respondents to give some context
    (see Appendix I for the prompt). The same scale was used for the
    answer's modalities than for the above question.

-   **Future intentions to use FP**

    This variable captures future intentions of participants to use
    contraceptives in the context of family planning. This was evaluated
    using a 4 point likert scale, similarly to the two variables
    described above. To conduct a logistical regression using this
    variable as outcome, this variable has been binarized as follows:

    -   "Some intentions": participants who answered "Strongly likely"
        or "Likely" to the "Future intentions to use FP" variable.

    -   "No intention": participants who answered "Unlikely" or
        "Strongly unlikely" to the "Future intentions to use FP"
        variable.

-   **Ever used**

    This variable was created to capture whether respondent ever used
    contraceptives throughout their life. It is build from two
    questions, one asking about participants current use of
    contraceptives, and another one asking respondents about their past
    use. This latter question was only asked to respondents who reported
    not using any contraceptive method currently. Participants who
    reported either using a contraceptive method currently, or used some
    in the past, where assigned a "Yes" to the ever used variable, while
    other were assigned a "No".

-   **Economic status**

    Economic status has been built from two variables, one asking about
    respondent's monthly income, and in case respondents did not now
    their monthly income, they were asked about their monthly spending.
    The final economic status variable includes three modalities: low
    (monthly income \< KSH 23,670 or monthly spending \< KSH 4,999),
    medium (monthly income above KSH 23,670 but less than KSH 200,000,
    or monthly spending between KSH 5,000 and 49,999) and high (monthly
    income equal or above KSH 200,000, or monthly spending equal or
    above 50,000).

-   **Knowledge of Kala-azar**

    Knoweldge related to Kala-azar has been built from different
    questions related to Kala-azar means of transmission and symptoms.
    It includes four modalities:

    -   "No knowledge": participants who reported not knowing Kala-azar
        as a disease.

    -   "Transmission knowledge": participants who know at least one
        mean of transmission of Kala-azar.

    -   "Symptom knowledge": participants who know at least one symptom
        related to Kala-azar.

    -   "Complete knowledge": participants who know both at least one
        symptoms and means of transmission associated with Kala-azar.

-   **Experience of Kala-azar**

    This variable classifies participants based on how much they have
    been in contact, or heard about, people sick with Kala-azar. It
    includes three modalities:

    -   "Seen/heard of sickness": participants who report having seen or
        heard about someone with Kala-azar infection.

    -   Seen/heard of death": participants who report having seen or
        heard about someone dying from Kala-azar infection.

    -   "No contract": participants who neither have seen/heard about
        someone infected or who died of Kala-azar.

-   **Feared treatment-related consequences**

    Number of consequences associated with Kala-azar treatment feared by
    participants (e.g. pain related to injections, prolonged hospital
    stay, anxiousness/depression). This variable includes two
    categories:

    -   "No/low concerns": participants who reported fearing not more
        than one treatment related consequence.

    -   "Moderate/high concerns": participants who expressed fearing
        more than 1 treatment related consequences.

-   **Knowledge of contraceptive methods**

    Number of contraceptive methods known by participants, which
    includes 3 modalities:

    -   "Low knowledge": participants who reported knowing between 0 and
        2 contraceptive methods.

    -   "Moderate knowledge": partcipants who reported knowing between 3
        and 4 contracptive methods.

    -   "High knowledge": participants who reported knowing more than 4
        contraceptive methods.

-   **Fear of side effects**

    This variable follows the framing tested and validated by Zan &
    Rossier (2024). It rests on two questions: one related to the number
    of side effects known by respondent and the number of these that are
    feared. The number of side effects feared is then divided by the
    number of side effect known, which yields a scale varying between 0
    and 1 (0 expressing no concerns, and 1 expressing the highest level
    of concern). Participants who reported knowing no side effects where
    considered as having no fear (i.e. a value of 0 for this variable).
    This continuous indicators has then been classified in three
    modalities for better understanding:

    -   "Low fear": participants whose responses yield a result ranging
        from 0 to 0.33.

    -   "Moderate fear": participants whose responses yield a result
        ranging from 0.34 to 0.66.

    -   "High fear": participants whose responses yield a results
        ranging from 0.67 to 1.

-   **Contraceptive use approval**

    This variable follows the framing tested and validated by Zan &
    Rossier (2024). It is build on three questions related to the
    approval of the use of contraceptive within a couple in different
    context: to prevent birth, to space birth and to limit birth. These
    questions are evaluated on a 5 points likert scale which are
    assigned a value between 1 and 5 (with "Strongly agree" being
    assigned a value of 5 and "Strongly disagree" a value of 1). The
    value of the three questions have then been summed up for each
    participant, and thus yields a value varying between 3 and 15. This
    continuous indicator has then been framed in the following three
    modalities:

    -   "Low approval": participants whose answers yield a value varying
        between 3 and 6.

    -   "Medium approval": participants whose answers yield a value
        varying between 7 and 10.

    -   "High approval": participants whose answers yield a value above
        10.

-   **Contraceptive use agency**

    This variable follows the framing tested and validated by Zan &
    Rossier (2024). It has been framed following the same process than
    for the contraceptive use approval variable described above (i.e.
    the values associated to the likert scale modalities are summed up
    to form a single value) - however, this variable rests on 5
    questions related to contraceptive use agency, yielding a single
    value varying between 5 and 25. This continuous indicator has then
    been framed in the following three modalities:

    -   "Low agency": participants whose answers yield a value varying
        between 5 and 13.

    -   "Medium agency": participants whose answers yield a value
        varying between 14 and 19.

    -   "High approval": participants whose answers yield a value above
        19.

-   **Decision maker**

    This variable has been framed from two variables asking about the
    ability of respondents to decide using contraceptives for
    themselves. The first one ask about whose taking such a decision
    (e.g. herself, their partner, herself and their partner jointly).
    Respondents who answered "Respondent and partner jointly" where then
    ask whose opinion between themselves and their partners matters the
    most in such decisions. The final decision maker variable includes
    two modalities:

    -   "Respondent": participants who expressed that they are taking
        decision related to their own use of contraceptives themselves,
        or, if respondents expressed that this decision is taken
        jointly, that their own opinion matters more or equally to the
        one of their partner concerning this topic.

    -   "Husband or someone else": participants who reported that their
        partner of someone else is taking decision related to their own
        use of contraceptives, or, if respondents expressed that this
        decision is taken jointly, that the opinion of their partner
        matters more than their own concerning this topic.

# 3. Results

## Descriptive analysis

A total of 401 WOCBP were interviewed from Kacheliba and Sigor District
Hospitals. Socio-demographic characteristics are found in Table 1.

##### Table 1

```{r Table 1, echo=FALSE}
htmltools::tags$iframe(
  src = here::here("C:/Kenya_anaylsis_Rproject/Project_outputs/outputs_touse/tables/final_cat_var_table.html"),
  width = "100%",
  height = "600px",
  style = "border:none;"
  )
```

#### *Socio-demographic variables*

Concerning the age of participants, close to half of the sample is
between 18 and 24 years old, and the teenager category is the least
represented modality within the sample (21 participants, 5.2%). Most
participants gave birth to 1 or 2 children (n = 166, 41.4%), which tends
to be quite in line with the age distribution of the sample (i.e. 43.1%
of participants have between 18 and 24 years old).

A majority of participants are living in rural areas (92.3%), sedentary
(65.8%), and of Christian confession (96.8%). Furthermore, Almost half
participants did not complete any level of formal education (48.1%) ,
and a quarter did complete primary education (25.7%). Most participants
were considered as having a low economic status (81%) or medium economic
status (18.2%).

#### *Knowledge and perceptions of leishmaniasis*

Knowledge of leishmaniasis was quite high among participants, with 33.4%
of the sample knowing both at least a mean of transmission and one
symptom, and 29.4% of participants who either never heard about
leishmaniasis, or did not know any symptom nor means of transmission.
Most participants had some experience with leishmaniasis, with 56.6% of
the sample having heard or seen someone sick from leishmaniasis, and
21.7% heard or seen someone die from this disease. Furhtermore,
participants showed relatively low concerns about consequences of
leishmaniasis treatment, with above 70% of participants evaluated as
having no or low concerns. Pain at the injection site (31.7%) and
prolonged hospital stay (26.9%) were the most frequent concerns
mentioned by participants.

#### *Contraceptive use variables*

Approximately a third of participants reported using contraceptives when
interviewed (34.7%), while this rate jumps to 55.9% when reporting about
ever usage of contraceptives. Figure 1 describes the kind of
contraceptives methods reported to be used by participants who ever used
contraceptives. Three quarter of participants reported being the
decision makers related their use of contraceptives (75.1%), and 10.7%
reported to have been pressured to give birth at some point in the past.

##### Figure 1

```{r Figure 1, echo=FALSE}
N_ever_user <- sum(!is.na(graphs_dataset$use_type))

univar_use_type <- 
  graphs_dataset %>% 
  mutate(use_type = case_when(use_type == 1 ~ "Female sterilization",
                              use_type == 2 ~ "Male sterilization",
                              use_type == 3 ~ "Intra-uterine device (IUD)",
                              use_type == 4 ~ "Injectables",
                              use_type == 5 ~ "Implants",
                              use_type == 6 ~ "Pills",
                              use_type == 7 ~ "Condom",
                              use_type == 8 ~ "Female condom",
                              use_type == 9 ~ "Emergency contraception",
                              use_type == 10 ~ "Standard days method",
                              use_type == 11 ~ "Lactational amenorrhea method",
                              use_type == 12 ~ "Other")) %>%
  count(use_type) %>% 
  mutate(percentage = n / N_ever_user * 100,
         tooltip_label = paste(use_type, " (n = ", n, "): ", round(percentage, 2), "%", sep = "")) %>%
  filter(!is.na(use_type)) %>% 
  ggplot(aes(x = reorder(use_type, n), y = percentage, fill = use_type, text = tooltip_label)) +
  geom_col() +
  labs(title = "Type of modern contraceptive methods used among ever-users",
       x = "Type of modern contraceptive method",
       y = "Percentage") +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plotly_univar_use_type <- ggplotly(univar_use_type, tooltip = "text")

plotly_univar_use_type
```

More than three quarter of participants reported being strongly likely
(48.1%) or likely (30.7%) to use contraceptives for family planning in
the future. In the context of clinical trials, 81% of participants
reported being either strongly likely (50.1%) or likely (30.9%) to
choose a treatment regiment that requires to use a contraceptive method
during treatment over one that do not require contraceptive use,
provided that the former treatment is less disruptive (in terms of
number of injections, length of hospital stay and side effects) - this
rate jumps to 85.1% (strongly likely: 58.4%, likely: 26.7%) in the
context of standard treatment (i.e. outside clinical trials).

#### *Psycho-social variables*

Participants mostly displayed moderate contraceptive knowledge (72.6%),
and had high fear of side effects(45.4%). The kind of feared side
effects are displayed in Figure 2. Concerning contraceptive use
approval, three quarter of the sample displayed high approval (75. 8%) -
however, 45.6% of participants described a high agency regarding
contraceptive use, and 42.4% were categorized as having moderate agency.

##### Figure 2

```{r Figure 2, echo=FALSE}
univar_fear_side_eff <- 
  graphs_dataset %>%
  # Split combined_reasons into individual reasons, while retaining the original structure
  mutate(f_side_eff_combined_list = str_split(f_side_eff_combined, ",\\s*")) %>%
  unnest(f_side_eff_combined_list) %>%  # Create a column of individual reasons
  filter(!is.na(f_side_eff_combined_list)) %>%  
  count(f_side_eff_combined_list) %>% 
  mutate(percentage = n / sum(n) * 100, # nrow(num_data_renamed) inserted manually to avoid counting 
         # from split reasons rather than the total sample size
         tooltip = paste(f_side_eff_combined_list, " (n = ", n, ") ", round(percentage, 2), "%", sep = "")) %>% 
  ggplot(aes(x = reorder(f_side_eff_combined_list, n), y = n, fill = f_side_eff_combined_list, text = tooltip)) +
  geom_col() +
  labs(x = "Side effects feared", y = "Count", title = "Counts of side effects feared by respondents") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none")

plotly_univar_fear_side_eff <- ggplotly(univar_fear_side_eff, tooltip = "text")

plotly_univar_fear_side_eff
```

## Binary analysis: difference in proportions

In order to establish whether a significant difference exists between
participants reporting future intentions to use contraceptives for
family planning and intentions to use some contraceptive methods in the
context of standard treatment for leishmaniasis, a McNemar test was
applied these two proportions. To do so, both of these variables were
binarized - as shown in Table 2, both were originally recorded using a 4
point likert-scale. The results of this test are found in Table 3.

##### Table 3

```{r echo=FALSE}
htmltools::tags$iframe(
  src = here::here(
    "C:/Kenya_anaylsis_Rproject/Project_outputs/outputs_touse/tables/final_mcnemar_gt_treat.html"),
  width = "100%",
  height = "325px",
  style = "border:none;"
  )
```

## Multivariate analysis

Table 4-6 display three logistical regression models that differ
structurally because of their respective outcomes: Table 4 includes
intentions to use contraceptives alongside medical treatment as outcome,
Table 5 includes future intentions to use contraceptives as outcome,
while Table 6 includes ever use of contraceptive as outcome. These
models have been set up in order to compare the difference in predictors
for each outcome, thereby allowing to identify difference in factors
influencing contraceptives use intentions (and behaviour for ever use)
in these different contexts (for medical reasons and for family planning
purpose).

##### Table 4

```{r echo=FALSE}
htmltools::tags$iframe(
  src = here::here("C:/Kenya_anaylsis_Rproject/Project_outputs/outputs_touse/tables/finalfit_tables/gt_finalfit_treatment.html"),
  width = "100%",
  height = "600px",
  style = "border:none;"
  )
```

Concerning the psychosocial predictors included in the model described
in Table 4, all categories related to the approval of contraceptive use
and agency with regard to contraceptive use are highly negatively
associated with intention to use contraceptives alongside medical
treatment (all p-values under 0.01) - fear of side effects seems to be
positively associated with the outcome (although in a less significant
manner as shown by the p-value), although only when observing the
moderate category. Similarly, ever use of contraceptives is also
positively associated with the outcome. However, it is important to note
that the ability of participants to decide using contraceptives for
herself is not associated with intention to use contraceptives alongside
medical treatment, as shown by the odds ration and p-value related to
this variable. Furthermore, knowledge of contraceptive shows no
association with intention to use contraceptives alongside medical
treatment.

For demographic predictors, age and number of children are both highly
associated with intention to use contraceptive alongside medical
treatment, although in different directions (negatively and positively,
respectively). Somehow surprisingly, higher economic status tends to be
negatively associated with the outcome, when one could expect it to be
positively associated.

Finally, disease related predictors also show some interesting results:
as the number of concerns related to treatment consequences, intentions
to use contraceptives for medical reasons increase. This is not
surprising, since the treatment regiment for leishmaniasis which does
involve using contraceptives present more negative consequences than the
regiment requiring to use contraceptives (a precision which was stated
in the prompt of the question related to intention to use contraceptives
alongside medical treatment). Finally, being knowledgeable about the
symptoms of leishmaniasis, in comparison to being knowledgeable about
both symptoms and means of transmission of leishmaniasis, is positively
associated with the outcome.

##### Table 5

```{r Table 5, echo=FALSE}
htmltools::tags$iframe(
  src = here::here("C:/Kenya_anaylsis_Rproject/Project_outputs/outputs_touse/tables/finalfit_tables/gt_finalfit_FP.html"),
  width = "100%",
  height = "600px",
  style = "border:none;"
  )
```

##### Table 6

```{r Table 6, echo=FALSE}
htmltools::tags$iframe(
  src = here::here("C:/Kenya_anaylsis_Rproject/Project_outputs/outputs_touse/tables/finalfit_tables/gt_finalfit_everused.html"),
  width = "100%",
  height = "600px",
  style = "border:none;"
  )
```

What can be observed when comparing these three models is that first

# Discussion

-   difference between the models

    -   agency and decision maker varies (decision_maker not significant
        in medical context but significant for FP - agency strongly
        associated in medical context but not really in FP context)

    -   age and n_birth are highly significant in medical context but
        not so much for family planning

    -   "Classic" predictors of contraceptive use are not so much
        associated with the medical context, in comparison to FP
        context.

-   why age (and n_birth) is significant only in medical context

-   issue with the fear of SE variable

-   8% of the sample fears issues related to fertility when using FP

-   drinking unboiled milk have been often reported as a cause of
    leishmaniasis

What can be taken out of this table is first that a strong and
significant difference exists between the proportion compared (McNemar
Chi-Squared = 8.6, p-value = 0.0032). The absolute difference further
confirm the direction of this difference, with future intentions to use
contraceptive for family planning being 6.5% lower than intention to use
contraceptives alongside leishmaniasis treatment (CI = -10.9%, - 2.2%).
This tends to indicate that decision related to using contraceptives in
these two different contexts are actually taken differently, and
identifying predictors related to these two phenomenon are further
inquired in the subsequent section.
